A tutorial of Bayesian beta regressions with brms in R

### 2024-08-02 manuscript/manuscript.qmd
edit manuscript

@@

```{r}
library(tidyverse)
theme_set(theme_light())

load("data/coretta2018/ita_egg.rda")

ita_egg <- ita_egg |> 
  filter(voi_clo_prop > 0, voi_clo_prop < 1) |> 
  mutate(
    vowel = factor(vowel, levels = c("i", "e", "a", "o", "u")),
    c2 = factor(c2, levels = c("k", "t", "p"))
  )
```

```{r}
ita_egg
```

```{r}
#| label: fig-ita-egg

@@

```{r}
#| label: brms
#| message: false

library(brms)
```

```{r}
# Takes 3 minutes on MacBook Pro 2020, M1

voi_prop_bm <- brm(
  voi_clo_prop ~
    0 + vowel:c2 + speech_rate_c +
    (0 + vowel:c2 + speech_rate_c | speaker),
  family = Beta,
  data = ita_egg,
  cores = 4,
  seed = 3749,
  file = "data/cache/voi_prop_bm"
)
```

```{r}
summary(voi_prop_bm, prob = 0.8)
```

@@

comp_c2 <- avg_comparisons(voi_prop_bm, variables = list(c2 = "pairwise"), conf_level = 0.8, type = "prediction")
```

>>>>

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(brms)
```

@@

For the first case study, we will model the proportion of voicing within consonant closure. The measurements come from a data set of audio and electroglottographic (EGG) recordings of 19 speakers of Northwestern Italian. The participants read frame sentences which included target words of the form /CVCo/, where /C/ was either /k, t, p/ in all permutations and /V/ was either /i, e, a, ɔ, u/ (two resulting words, /peto/ and /kako/ were excluded because they are profanities), for a total of 43 target words. There were 4 different frame sentence, with a total of 172 trials per participant (3,268 grand total). The actual observation count is 2,419, after removing speech errors, EGG measurement errors, and cases in which voicing ceased before the closure onset/after the closure offset of the second /C/.

The proportion of voicing during the closure of the second /C/ was calculated as the proportion of contiguous voicing duration after closure onset to total duration of closure. The following code chunk attaches the tidyverse packages (for reading and wrangling data) and loads the `ita_egg` tibble (data frame). The tibble is filtered so as to remove voicing proportions (`voi_clo_prop`) that are smaller than 0 and greater than 1. The variables `vowel` and `c2` are converted to factors to specify the order of the levels.

```{r}
#| label: read-ita-egg

# attach tidyverse and set light theme for plots
library(tidyverse)
theme_set(theme_light())

# load tibble
load("data/coretta2018/ita_egg.rda")

# filter and mutate data
ita_egg <- ita_egg |> 
  filter(voi_clo_prop > 0, voi_clo_prop < 1) |> 
  mutate(
    vowel = factor(vowel, levels = c("i", "e", "a", "o", "u")),
    c2 = factor(c2, levels = c("k", "t", "p"))
  )
```

Here is what the tibble looks like.

```{r}
#| label: ita-egg

ita_egg
```

@fig-ita-egg shows the raw voicing duration proportion values split by vowel /i, e, a, ɔ, u/ and second consonant /k, t, p/ in the /CVCo/ target words. The plot suggests that, on average, the voicing proportion is slightly lower with /k/ than with /t, p/. Moreover, there is greater variability between vowels in /t, p/ than in /k/. We will use a beta regression model to assess these patterns. ["expectations" XXX]

```{r}
#| label: fig-ita-egg
#| fig-cap: "Proportion of voicing during the closure of the second consonant in /CVCo/ words by vowel and the second consonant."
#| echo: false

@@

We will use brms to fit Bayesian beta regressions. [XXX why Bayesian]. 
The model will have voicing proportion as the outcome variable and the following terms: vowel (/i, e, a, ɔ, u/) and second consonant C2 (/k, t, p/) in interaction, centred speech rate, a varying intercept by speaker and varying slopes of vowel/consonant in interaction and centred speech rate (number of syllables per second). The categorical predictors vowel and C2 were coded using indexing (in R, this corresponds to suppressing the intercept with the `0 +` syntax; this type of predictor coding does not impact the model predictions, but it makes it easier to specify priors when users do so). For pedagogical reasons of simplicity, the model will use the default priors, but note that in real data analyses contexts, priors should be specified by the user. I refer the readers to XXX.

```{r}
#| message: false

# attach brms
library(brms)

# fit the model
# Takes 3 minutes on MacBook Pro 2020, M1
voi_prop_bm <- brm(
  # model formula
  voi_clo_prop ~
    # constant terms
    0 + vowel:c2 + speech_rate_c +
    # varying terms
    (0 + vowel:c2 + speech_rate_c | speaker),
  # uses the beta family for the outcome
  family = Beta,
  data = ita_egg,
  cores = 4,
  seed = 3749,
  file = "data/cache/voi_prop_bm"
)
```

The `summary()` function prints the full model summary. For conciseness, we will use the `fixef()` function which prints the regression coefficients of the constant (aka fixed) terms. The full summary with an explanation of each part can be found in XXX. For each coefficient in the model, `fixef()` prints the name of the coefficient, the mean estimate, the estimate error and the lower and upper limits of a Bayesian Credible interval (CrI). Here, we print an 80% CrI.^[There is nothing special about 95% within Bayesian inference and instead experts recommend to check and report a variety of CrIs. For simplicity, I will focus on 80% CrIs in this case study.] In Bayesian inference, an 80% CrI indicates the range of values within which the true estimate falls at 80% probability or confidence. I round all values to the nearest 2 digits for clarity.

```{r}
round(
  fixef(voi_prop_bm, prob = c(0.1, 0.9)),
  digits = 2
)
```

The coefficients of a beta regression are estimated on the log-odds scale, as in Bernoulli/binomial (aka logistic) regressions. From the summary, we see that speech rate (number of syllables per second) has a positive effect on voicing proportion: the 80% CrI is between 0.01 and 0.15 log-odds [$\beta$ = 0.8, SD = 0.06]. Log-odds can be converted to odd ratios by exponentiating the value: 0.01-0.15 log odds correspond to an odd ratio of 1.01 to 1.16, or as percentages, to an increase of voicing of 1 to 16% for every increase of one syllable per second. Since this is an 80% CrI, we can be 80% confident that the true effect of speech rate is between 1-16% increase of voicing proportion.

Turning now to the coefficients for vowel and C2, given the indexing approach of coding these variables the model summary and the output of `fixef()` reports the *predictions* in log-odds for each combination of vowel and C2, rather than differences between levels. The CrIs of the vowel/C2 coefficients span all negative log-odds values: these correspond to proportions that are lower than 0.5 (which is 0 in log-odds). This matches the general trends in the raw data, which we plotted in @fig-ita-egg.

Next, we will plot the predicted proportions of each vowel/C2 combination at mean speech rate (i.e. centred speech rate = 0) and then calculate the average pair-wise difference in voicing proportion between /k, t, p/. Finally, we will assess whether there is greater between-vowel variability in /t, p/ relative to /k/.

Before being able to plot the predictions, it's important to get familiar with the so called posterior draws. [Bayesian MCMC XXX]. Posterior draws can be conveniently obtained with the `as_draws_df()`. For the moment, we can extract only the draws of the constant regression coefficients (model variables starting with `b_`). To check which coefficients are available in a model, use `get_variables()` from the tidybayes package. `as_draws_df()` returns a tibble where each column contains the drawn values of a coefficient. The probability distribution of the drawn values of each coefficient is the posterior probability distribution of that coefficient. Note that, due to the indexing coding of vowel and C2, all coefficient except `b_speech_rate_c` are *predicted log-odds* for each vowel/C2 combination (the drawn values for `b_speech_rate_c` are drawn *differences* in log-odds for each unit increase of speech rate). The drawn values are in log-odds, but we can convert them to proportions with `plogis()` (we will do this when plotting below).

```{r}
#| label: voi-prop-bm-draws

# extract only coefficient variables starting with "b_"
voi_prop_bm_draws <- as_draws_df(voi_prop_bm, variable = "^b_", regex = TRUE)
voi_prop_bm_draws
```

We can now wrangle this tibble and plot the posterior distributions for each vowel/C2 combination.

```{r}
voi_prop_bm_draws_long <- voi_prop_bm_draws |> 
  select(-b_speech_rate_c) |> 
  pivot_longer(`b_voweli:c2k`:`b_vowelu:c2p`, names_to = "coeff") |> 
  separate(coeff, into = c("type", "vowel", "c2"))
voi_prop_bm_draws_long
```

For plotting, we can use ggplot2 statistics from the ggdist package. `stat_halfeye()` plots the density of the posterior probability (in grey), its median (point) and CrIs (lines). Let's use 60 and 80% CrIs and transform the log-odds values to proportions with `plogis()`.

```{r}
#| label: fig-voi-prop-bm-1
# attach ggdist package
library(ggdist)

voi_prop_bm_draws_long |> 
  ggplot(aes(plogis(value), vowel)) +
  stat_halfeye(.width = c(0.6, 0.8)) +
  facet_grid(rows = vars(c2))
```

What if we want to plot the average predicted voicing proportion for the three consonants /k, t, p/? One approach is to take the mean across vowels within each consonant and the resulting posterior distribution is the predicted posterior distribution of voicing proportion for each consonant, assuming an "average" vowel.

```{r}
voi_prop_bm_draws_long_c2 <- voi_prop_bm_draws_long |> 
  group_by(.draw, c2) |> 
  summarise(
    value_mean = mean(value), .groups = "drop"
  )

voi_prop_bm_draws_long_c2 |> 
  ggplot(aes(plogis(value_mean), c2)) +
  stat_halfeye(.width = c(0.6, 0.8))
```

Based on the posterior distributions of the mean voicing proportion by consonant, /p/ has a somewhat higher voicing proportion than /k/ and /t/. The real question is: how higher? We can quantify this by taking the difference of the drawn values for /p/ and those for /t, k/. Since we want to compare /t, k/ to /p/, we should first average the draws of /t, k/ and then take the difference of the averaged draws and the draws of /p/.

```{r}
voi_prop_bm_diff <- voi_prop_bm_draws_long_c2 |> 
  pivot_wider(names_from = c2, values_from = value_mean) |> 
  mutate(
    c2tk = mean(c(c2k, c2t)),
    c2p_tk_diff = c2p - c2tk
  )

voi_prop_bm_diff |> 
  ggplot(aes(c2p_tk_diff)) +
  stat_halfeye(.width = c(0.6, 0.8, 0.9)) +
  geom_vline(xintercept = 0)
```

Once we have the posterior difference, we can obtain CrIs of that using `quantile2()` from the posterior package. Beware that the values of the difference are in log-odds! We can convert these into odd ratios with `exp()`.

```{r}
library(posterior)

voi_prop_bm_diff |> 
  reframe(
    # 90% CrI
    q90 = quantile2(c2p_tk_diff, probs = c(0.05, 0.95)),
    # 80% CrI
    q80 = quantile2(c2p_tk_diff, probs = c(0.1, 0.9)),
    # 60% CrI
    q60 = quantile2(c2p_tk_diff, probs = c(0.2, 0.8)),
  ) |> 
  # exponentiate to get odd ratios and round
  mutate(across(everything(), ~round(exp(.x), 2)))
```

So, based on the model and data, there is a 90% probability that the voicing proportion in /p/ is 1.07-1.38 times longer (or 7-38% increase) than in /t, k/. At 80% confidence, the change ratio is 1.10-1.34 (or 10-34% increase) while at 60% confidence is 1.14-1.30 (14-30% increase). In other words we can be quite confident that the voicing proportion in /p/ is longer than in /t, k/ and that the increase is less than 35%.

@@

avg_comparisons(voi_prop_bm, variables = list(c2 = "pairwise"), conf_level = 0.8)
avg_predictions(voi_prop_bm, variables = "vowel", conf_level = 0.8)
```

## Case study 2: pre-nasalisation in Greek
